An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-02-06 icml2017/paper.tex
Updated the paper to have lots of tables. I think now we can decide which sigma, lambda, etc. to set for the correction distribution
More updates to paper.
I added in a reference that a reviewer suggested. OK now focus on the BIDMach implementation...

Many state-of-the-art machine learning methods, and deep learning in particular,
are based on minibatch updates (such as SGD) to a model.  Minibatch updates
produce many improvements to the model for each pass over the dataset, and have
high sample efficiency.  In contrast, conventional M-H requires calculations
over the full dataset to produce a new sample.  Recent results
from~\cite{cutting_mh_2014} and~\cite{icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets of the full dataset.
The amount of data consumed for each test varies significantly from one
minibatch to the next. By contrast,~\cite{conf/uai/MaclaurinA14,TallData15}
perform exact tests but require a lower bound on the parameter distribution across
its domain.  The amount of data reduction depends on the accuracy of this bound,
and such bounds are only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test} and requires less data per
iteration than prior work. We use a Barker test function~\cite{Barker65}, which
makes our test naturally error tolerant. The idea of using a noise-tolerant
Barker's test function was suggested but not explored empirically
in~\cite{TallData15} section 6.3. But the asymptotic test statistic CDF and the
Barker function are different, which leads to fixed errors for the approach
in~\cite{TallData15}. Here, we show that the difference between the
distributions can be corrected with an additive random variable. This leads to a
test which is fast, and whose error can be made arbitrarily small.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$,
we can write
\begin{equation}\label{eq:relationship}
    \Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is
approximately normal with variance $\sigma^2(\Delta^*)$.  But to perform the
test in Equation~(\ref{eq:the_exact_test}) we want $\Delta + X$ for a logistic
random variable $X$ (call it $X_{\rm log}$ from now on). In~\cite{TallData15} it
was proposed to use $\Delta^*$ in a Barker test, and tolerate the fixed
error between the logistic and normal distributions.

@@

\textbf{Daniel: need to revise this to take into account reviewer suggestion,
with future work ideas. This will add some 2016 references so that'll be good
(makes it current)}

@@

% Daniel: moved to appendix so we can expand it. Originally we had only two
% lines, and it looked pretty awkward. So let's add more. Note: this also
% depends on whether we use [-10,10] for the X_c range, or [-20,20], etc...
\begin{table}[t]
\caption{Error ($L_\infty$) in $X_{\rm norm}+X_{\rm corr}$ versus $X_{\rm log}$.}
\label{tab:xcorr}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{l l l l}
\hline
\abovespace\belowspace
${\bf N}$ & ${\bf \sigma}$ & ${\bf \lambda}$ & ${\bf L_{\infty}}$ {\bf error} \\
\hline
\abovespace
4000 & 1.0 & 100    & 1.9e-3  \\
4000 & 1.0 & 10     & 8.9e-4  \\
4000 & 1.0 & 1      & 1.6e-3  \\
4000 & 1.0 & 0.1    & 2.8e-3  \\
4000 & 1.0 & 0.01   & 5.2e-3  \\
4000 & 1.0 & 0.001  & 9.9e-3  \\
2000 & 1.0 & 10     & 1.3e-3  \\
8000 & 1.0 & 10     & 1.3e-3  \\
\belowspace
4000 & 1.0 & 10 & 6.6e-4 \\
\hline
\end{tabular}
\quad
\begin{tabular}{l l l l}
\hline
\abovespace\belowspace
${\bf N}$ & ${\bf \sigma}$ & ${\bf \lambda}$ & ${\bf L_{\infty}}$ {\bf error} \\
\hline
\abovespace
4000 & 0.9 & 1 & 1.0e-4  \\
\belowspace
4000 & 0.8 & 0.03 & 5.0e-6 \\
\hline
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

>>>>

Many state-of-the-art machine learning methods, and deep learning in particular,
are based on minibatch updates (such as SGD) to a model.  Minibatch updates
produce many improvements to the model for each pass over the dataset, and have
high sample efficiency.  In contrast, conventional M-H requires calculations
over the full dataset to produce a new sample.  Recent results
from~\cite{cutting_mh_2014} and~\cite{icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets of the full dataset.
The amount of data consumed for each test varies significantly from one
minibatch to the next. By contrast,~\cite{conf/uai/MaclaurinA14,TallData16}
perform exact tests but require a lower bound on the parameter distribution across
its domain.  The amount of data reduction depends on the accuracy of this bound,
and such bounds are only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test} and requires less data per
iteration than prior work. We use a Barker test function~\cite{Barker65}, which
makes our test naturally error tolerant. The idea of using a noise-tolerant
Barker's test function was suggested but not explored empirically
in~\cite{TallData16} section 6.3. But the asymptotic test statistic CDF and the
Barker function are different, which leads to fixed errors for the approach
in~\cite{TallData16}. Here, we show that the difference between the
distributions can be corrected with an additive random variable. This leads to a
test which is fast, and whose error can be made arbitrarily small.

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$,
we can write
\begin{equation}\label{eq:relationship}
    \Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is
approximately normal with variance $\sigma^2(\Delta^*)$.  But to perform the
test in Equation~(\ref{eq:the_exact_test}) we want $\Delta + X$ for a logistic
random variable $X$ (call it $X_{\rm log}$ from now on). In~\cite{TallData16} it
was proposed to use $\Delta^*$ in a Barker test, and tolerate the fixed
error between the logistic and normal distributions.

@@

\textbf{Daniel: need to revise this to take into account reviewer suggestion,
with future work ideas. One reviewer suggested
using~\cite{DBLP:conf/icml/ChenG16}.}

@@

% Daniel: moved to appendix so we can expand it. Originally we had only two
% lines, and it looked pretty awkward. So let's add more. Note: this also
% depends on whether we use [-10,10] for the X_c range, or [-20,20], etc...
\begin{table}[t]
\caption{Errors ($L_\infty$) in $X_{\rm norm}+X_{\rm corr}$ versus $X_{\rm
log}$, with $N=4000$ (top row) and $N=2000$ (bottom row).}
\label{tab:xcorr}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
%% ----------------------- START OF TABLE
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=4000$ & $\sigma=1.1$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & \textbf{4.3e-3}  \\
10     & 6.0e-3  \\
1      & 1.0e-2  \\
0.1    & 1.2e-2  \\
\belowspace
0.01   & 3.5e-2  \\
\hline
\end{tabular}
\quad %% ----------------------- START OF TABLE
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=4000$ & $\sigma=1.0$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & 1.9e-3  \\
10     & \textbf{8.9e-4}  \\
1      & 1.6e-3  \\
0.1    & 2.8e-3  \\
\belowspace
0.01   & 5.2e-3  \\
\hline
\end{tabular}
\quad %% ----------------------- START OF TABLE
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=4000$ & $\sigma=0.9$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & 1.2e-3  \\
10     & 2.6e-4  \\
1      & \textbf{1.0e-4}  \\
0.1    & 2.0e-4  \\
\belowspace
0.01   & 3.9e-4  \\
\hline
\end{tabular}
\quad %% ----------------------- START OF TABLE
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=4000$ & $\sigma=0.8$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & 8.3e-4  \\
10     & 1.3e-4  \\
1      & 2.5e-5  \\
0.1    & \textbf{6.7e-6}  \\
\belowspace
0.01   & 7.4e-6  \\
\hline
\end{tabular}
%% ---- NEW ROW ----
\vskip 0.2in %% OH this works!
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=2000$ & $\sigma=1.1$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & 6.8e-3  \\
10     & \textbf{4.6e-3}  \\
1      & 7.5e-3  \\
0.1    & 1.3e-2  \\
\belowspace
0.01   & 2.4e-2  \\
\hline
\end{tabular}
\quad %% ----------------------- START OF TABLE
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=2000$ & $\sigma=1.0$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & 4.4e-3  \\
10     & 1.3e-3  \\
1      & \textbf{1.1e-3}  \\
0.1    & 2.0e-3  \\
\belowspace
0.01   & 3.6e-3  \\
\hline
\end{tabular}
\quad %% ----------------------- START OF TABLE
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=2000$ & $\sigma=0.9$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & 3.3e-3  \\
10     & 6.4e-4  \\
1      & 1.6e-4  \\
0.1    & \textbf{1.3e-4}  \\
\belowspace
0.01   & 2.7e-4  \\
\hline
\end{tabular}
\quad %% ----------------------- START OF TABLE
\begin{tabular}{l l}
\hline
\abovespace\belowspace
$N=2000$ & $\sigma=0.8$ \\
\hline
\abovespace\belowspace
$\lambda$ & $L_{\infty}$ error \\
\hline
\abovespace
100    & 2.6e-3  \\
10     & 4.0e-4  \\
1      & 6.7e-5  \\
0.1    & 1.4e-5  \\
\belowspace
0.01   & \textbf{5.0e-6}  \\
\hline
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}
